---
title: "Assemble dataset"
author: "Maxime Fraser Franco"
date: "January 9, 2024"
toc: true
format:
  html
#execute:
#  warning: false
---

# Import the data from repositories

## Original dataset

Here, I import the original dataset used in Fraser Franco et al. (2024). The data is hosted on a GitHub repository that you can access [here](https://github.com/quantitative-ecologist/experience-hunting-tactics).


```{r}
github_id <- "https://raw.githubusercontent.com/quantitative-ecologist/"
repo <- "experience-hunting-tactics"
file <- "/main/data/FraserFrancoetal2023-data.csv"
dat <- read.csv(
  file.path(github_id, repo, file)
)
```

Check the data structure

```{r}
head(dat)
```

## Dataset generated from Bayesian hierarchical model

Here, we import the predicted values generated from the Bayesian multivariate double-hierarchical model. This allows us to assess whether predators were specialist or flexible at different experience levels. To do so, we import the `.rds` files which are hosted on this [OSF repository](https://osf.io/hdv38/)